---
title: "Race and the Prison/Jail System"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(styler)
library(maps)
library(tmap)
library(knitr)
library(reshape2)
```

Problem:
Through the data collected by Vera Institute, I want to see if there is a racial bias towards black people in the
prison and jail system. 


```{r}
#Extracting data
urlfile = "https://raw.githubusercontent.com/vera-institute/incarceration-trends/master/incarceration_trends.csv"

#Naming the dataset
incar_trends <- read_csv(url(urlfile))

#Looking at the columns
incar_var <- colnames(incar_trends)

#Different regions
incar_regions <- unique(incar_trends$region)
```


Basic Look at the data set,
incar_pop looks at the change over time on the increase of the US population
and incar_jnp looks specifically at prisons and jails. 
```{r}
#Looking at the amount of people that are ages 15 to 64 from 1970 to 2010
incar_pop <- incar_trends %>%
  group_by(year) %>%
  summarize(total_pop_15to64 = sum(total_pop_15to64)) %>%
  filter(year >= '1990' & year <= '2010')

#Graph of incar_pop
pop_line <- ggplot(incar_pop, aes(year, total_pop_15to64)) + geom_line(color = 'darkgreen', size = 1.5) + labs(x = 'Year', y = 'Total Pop') + ggtitle ('Total population
from ages 15 to 64 over time')
pop_line

#Looking at the amount of people in prison and jail from 1990 to 2010
incar_jnp <- incar_trends %>%
  group_by(year) %>%
  summarize(total_jail_pop = sum(total_jail_pop,na.rm=TRUE), 
            total_prison_pop = sum(total_prison_pop,na.rm=TRUE)) %>%
  filter(year >= '1990' & year <= '2010')

#Graph of incar_jnp
jnp_lines <- ggplot(incar_jnp) + 
  geom_line(aes(x = year, y = total_jail_pop), size = 1.5, color = 'darkred') +
  geom_line(aes(x = year, y = total_prison_pop), size = 1.5, color = 'darkblue') +
  ylab('Total Pop') +
  xlab('Year') +
  ggtitle('Total pop in Jail (Red)  vs. Prison (Blue) from years 1990 to 2010') 
jnp_lines
```


There has been a sharp increase in terms of total population that can go to jail in the US. The prison and jail population also have an increase but for the most part there are way more people going to prison. At the start of the data, there are way more people in prison compared to jail but the increase for prison is sharper compared to jail and the jail population has occasional dips. 



```{r}
#Race over time in Jail
incar_jail_race <- incar_trends %>%
  filter(year >= '1985') %>%
  group_by(year) %>% 
   summarize(black_jail_pop = sum(black_jail_pop,na.rm = TRUE),
            white_jail_pop = sum(white_jail_pop, na.rm = TRUE),
            latinx_jail_pop = sum(latinx_jail_pop, na.rm = TRUE),
            aapi_jail_pop = sum(aapi_jail_pop,na.rm = TRUE),
            native_jail_pop = sum(native_jail_pop,na.rm = TRUE),
            other_race_jail_pop = sum(other_race_jail_pop,na.rm = TRUE))

#Graph of incar_jail_race
jail_line <-  ggplot(data = incar_jail_race, aes(x = year)) +
  geom_line(aes(y = black_jail_pop, colour = 'black_jail_pop')) +
  geom_line(aes(y = white_jail_pop, colour = 'white_jail_pop')) +
  geom_line(aes(y = latinx_jail_pop, colour = 'latinx_jail_pop')) +
  xlab('Years') +
  scale_colour_manual('',  
                      breaks = c('black_jail_pop','white_jail_pop','latinx_jail_pop'),
                      values = c('red','green','blue')) +
  scale_y_continuous('Total in Jail') +
  labs(title = 'Race in Jail overtime')
jail_line


#race over time in prison
incar_prison_race <- incar_trends %>%
  filter(year >= '1983' & year <= '2016') %>%
  group_by(year) %>%
  summarize(black_prison_pop = sum(black_prison_pop,na.rm = TRUE),
            white_prison_pop = sum(white_prison_pop, na.rm = TRUE),
            latinx_prison_pop = sum(latinx_prison_pop,na.rm = TRUE))

#Graph of incar_prison_race
prison_line <- ggplot(data = incar_prison_race, aes(x = year)) +
  geom_line(aes(y = black_prison_pop, colour = 'black_prison_pop')) +
  geom_line(aes(y = white_prison_pop, colour = 'white_prison_pop')) +
  geom_line(aes(y = latinx_prison_pop, colour = 'latinx_prison_pop')) +
  xlab('Years') + 
  scale_colour_manual('',  
                      breaks = c('black_prison_pop','white_prison_pop','latinx_prison_pop'),
                      values = c('red','green','blue')) +
  scale_y_continuous('Total in Prison') +
  labs(title = 'Race in prison overtime')
prison_line
```


For prison vs. jail in terms of race, there has consistently been more black people going to prison compared to white people and is mostly opposite for the jail population. One thing to take note is that the black and jail population in prison and jail are fairly close, especially comparing to latinx, but there are way more white people than black people in the US. 

```{r}
#Percent of each race in jail for the year 2016
jail_percent <- incar_trends %>%
  filter(year == '2016' ) %>% 
  group_by(region) %>%
  summarize(black_percent = sum(black_jail_pop,na.rm=TRUE)/sum(black_pop_15to64,na.rm = TRUE),
            white_percent = sum(white_jail_pop,na.rm=TRUE)/sum(white_pop_15to64,na.rm = TRUE),
            latinx_percent = sum(latinx_jail_pop,na.rm=TRUE)/sum(latinx_pop_15to64,na.rm = TRUE), 
            aapi_percent = sum(aapi_jail_pop,na.rm=TRUE)/sum(aapi_pop_15to64,na.rm=TRUE),
            native_percent = sum(native_jail_pop,na.rm=TRUE)/sum(native_pop_15to64,na.rm=TRUE)) 
jail_percent <- melt(jail_percent)

#Graph of jail_percent
jail_bar <-ggplot(jail_percent, aes(x=variable, y=value, fill=variable)) +
  geom_bar(stat='identity', position='dodge') + 
  scale_y_continuous('Percent',labels=scales::percent) + 
  xlab('Race') + 
  ylab('Percent') + 
  labs(fill='Race Percent', title = 'Percent of the total pop of 15to64 for each race that are in Jail')
jail_bar

#Percent of each race in prison for the year 2016
prison_percent <- incar_trends %>% 
  filter(year == '2016' ) %>%
  summarize(black_percent =   sum(black_prison_pop,na.rm  = TRUE)/sum(black_pop_15to64,na.rm=TRUE),
            white_percent =  sum(white_prison_pop,na.rm  = TRUE)/sum(white_pop_15to64,na.rm=TRUE),
            latinx_percent =  sum(white_prison_pop,na.rm  = TRUE)/sum(latinx_pop_15to64,na.rm=TRUE),
            aapi_percent =  sum(aapi_prison_pop,na.rm   = TRUE)/sum(aapi_pop_15to64,na.rm=TRUE),
            native_percent = sum(native_prison_pop,na.rm = TRUE)/sum(native_pop_15to64,na.rm=TRUE))
prison_percent <- melt(prison_percent)

prison_bar <- ggplot(prison_percent,aes(x = variable, y = value, fill = variable)) + 
  geom_bar(stat = 'identity', position = 'dodge') + 
  scale_y_continuous(labels=scales::percent) +
  xlab('Race') +
  ylab('Percent') +
  labs(fill = 'Race Percent', title = 'Percent of the total pop of 15to64 for each race that are in Prison')
prison_bar
```


Looking at the percentages it shows that there is a higher percentage of the black population going to prison and jail compared to white people. For prison, white people rank third in terms of the percentage of there race that are jail. 





